Stagnating State Funding for Higher Education and its Effect on Faculty at US Universities

### 2023-08-21 text/sections/00-abstract.tex
adjusted writing.

US public universities universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while their states have systematically reduced the funding to these universities relative to enrolment.
This paper connects the changes the faculty composition at US public universities to the fall in the states' funding for their public universities, using a shift-share approach to instrument for state funding and identify the causal effects of stagnating funding for higher education.
A fall in funding of \$1,000 per student leads to univeristies employing 6 more lecturers on average; relative to enrolment, a fall of $10$\% in funding per student decreases the number of assistant professors per student by $1.4$\% and full professors by $1.2$\%, yet increases the number of lecturers per student by $4.4$\%.
Dynamic estimates show that these effects linger multiple years after the initial funding shock.
Analysis of all faculty at Illinois public universities shows that incumbent professors' salaries, promotion rate, and quit rate are all are not affected by funding cuts over 2011-2021, implying that the composition change arose by limiting the hiring of new tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and exhibit one possible mechanism for deteriorating student outcomes at US public universities over this same time period.

>>>>

US public universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while their states have systematically reduced the funding to these universities relative to enrolment.
This paper connects the changes the faculty composition at US public universities to the fall in the states' funding for their public universities, using a shift-share approach to instrument for state funding and identify the causal effects of stagnating funding for higher education.
A fall in funding of \$1,000 per student leads to univeristies employing 6 more lecturers on average; relative to enrolment, a fall of $10$\% in funding per student decreases the number of assistant professors per student by $1.4$\% and full professors by $1.2$\%, yet increases the number of lecturers per student by $4.4$\%.
Dynamic estimates show that these effects linger multiple years after the initial funding shock.
Analysis of all faculty at Illinois public universities shows that incumbent professors' salaries, promotion rate, and quit rate are all are not affected by funding cuts over 2011-2021, implying that the composition change arose by limiting the hiring of new tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and exhibit one possible mechanism for deteriorating student outcomes at US public universities over this same time period.

### 2023-08-21 text/sections/01-introduction.tex
adjusted writing.

Public universities educate the majority of higher education students in the US, yet have experienced a stagnation in state funding over the last three decades .
At the same time, public university employment of lecturers dramatically increased at public universities, while enrolment rose dramatically.
Previous research has examined the effects of the decline in state funding on student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence of the effects of this funding decline on faculty.
I show that four-year, degree-granting public universities have experienced a systematic decline in state funding per student, and use a shift-share instrument to show that this fall led to a substitution away from tenure-track/tenured professors towards contingent lecturers.
Analysis of all public university faculty in the state of Illinois shows that incumbent faculty were relatively unaffected, in terms of wages, promotion and quit rates, implying that changes in faculty composition arose by disrupting public universities' ability to hire or replace their tenure-track/tenured faculty. 
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors, so that the stagnating state funding for public universities has implications for the wider structure of US higher education and research.

>>>>

Public universities educate the majority of higher education students in the US, yet have experienced a stagnation in state funding over the last three decades.
At the same time, use of lecturers dramatically increased, while enrolment rose dramatically.
Previous research has examined the effects of the decline in state funding on student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence of the effects of this funding decline on faculty.
I show that four-year, degree-granting public universities have experienced a systematic decline in state funding per student, and use a shift-share instrument to show that this fall led to a substitution away from tenure-track/tenured professors towards contingent lecturers.
Analysis of all public university faculty in the state of Illinois shows that incumbent faculty were relatively unaffected, in terms of wages, promotion and quit rates, implying that changes in faculty composition arose by disrupting public universities' ability to hire or replace their tenure-track/tenured faculty. 
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors, so that the stagnating state funding for public universities has implications for the wider structure of US higher education and research.
